---
title: "hw2-Gapminder Exploratory Data Analysis"
author: "Irfan Ainuddin"
date: "9/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("gapminder")
library("tidyverse")
```

## 1.1 - Loading and Looking
Load the Gapminder data .

```{r}
## Loading in the gapminder library which contains the data to be explored in this exercise.


## library("gapminder")

```

Load the tidyverse to work with data.
```{r}
## Loading in the complete tidyverse package (which has extra tools we wont use) to begin exploratory data analysis (EDA) of the gapminder datased loaded above.


## library("tidyverse")


```


Look at the structure of the gapminder data set.
```{r}

## Inspect the structure of the data with the str() function.

str(gapminder)

##  Display the first 6 values of the data set with the head() function.

head(gapminder)
```

The gapminder data set includes 6 variables (columns) with 1704 observations (rows).

The variables included are: country, continent, year, lifeExp, pop, gdpPercap.

country, and continent are factors (categorical variables), year and pop are integers (whole numbers) and lifeExp and gdpPercap are numbers (numbers with decimal places).


## 1.2 Goal Setting
The idea is to take a large data set with complexity and break it down into component parts. In this case, we have continents which contain countries which have a population with a life expectancy that was determined over some period of time.


I would like to be able to investigate the variation of lifeExp and population between different countries within a specific continent for the years in which the data is available.


## 1.3 Inspect the variation within the variables


```{r}
## Call the summary() function on the year variable to see the range of years data is available

summary(gapminder$year)

```

The years range from 1952 to 2007, 55 years total.

```{r}
## Call the table() function on the continent variable in the gapminder dataset to count the number of observations recorded for each continent.

table(gapminder$continent)

```

From the table we can see the number of observations for each country, Africa has the highest frequency (n = 624) and Oceania with the lowest frequencey (n = 24)

Since the goal is to be able to work with complex data and break it down into component parts, I will focus on the continent of Africa.

```{r}


```

```{r}



```


```{r}

```


```{r}


```


```{r}


```

```{r}

```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}

```


```{r}

```


```{r}

```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
