---
title: "hw2-Gapminder Exploratory Data Analysis"
author: "Irfan Ainuddin"
date: "9/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("gapminder")
library("tidyverse")
```

## 1.1 Loading and Looking
Load the Gapminder data .

```{r}
## Loading in the gapminder library which contains the data to be explored in this exercise.


## library("gapminder")

```

Load the tidyverse to work with data.
```{r}
## Loading in the complete tidyverse package (which has extra tools we wont use) to begin exploratory data analysis (EDA) of the gapminder datased loaded above.


## library("tidyverse")


```


Look at the structure of the gapminder data set.
```{r}

## Inspect the structure of the data with the str() function.

str(gapminder)

##  Display the first 6 values of the data set with the head() function.

head(gapminder)
```

The gapminder data set includes 6 variables (columns) with 1704 observations (rows).

The variables included are: country, continent, year, lifeExp, pop, gdpPercap.

country, and continent are factors (categorical variables), year and pop are integers (whole numbers) and lifeExp and gdpPercap are numbers (numbers with decimal places).


## 1.2 Goal Setting
The idea is to take a large data set with complexity and break it down into component parts. In this case, we have continents which contain countries which have a population with a life expectancy that was determined over some period of time.


I would like to be able to investigate the variation of lifeExp and population between different countries within a specific continent for the years in which the data is available.


## 1.3 Inspect the variation within the variables


```{r}
## Call the summary() function on the year variable to see the range of years data is available

summary(gapminder$year)
table(gapminder$year)

```

The years range from 1952 to 2007, however, the table shows us there are only 12 years of that range which data was collected. In this case, while the variable is an int, it sort of acts as a factor (categorical variable).

```{r}
## Call the summary() function on the pop variable 

summary(gapminder$pop)

## Call the hist() function to view the population distributions
hist(gapminder$pop)

```

The range in population is quite large from 6.001e+04 to 1.319e+09, considering sthe scale range of values suggests exponential changes in population. We may need to apply a logarithmic scale to visualize population data in a linear relationship with other variables. A quick look at the histogram shows the frequency of observations, since we know populations belong to each country

```{r}
## Call the summary() function on the lifeExp variable 

summary(gapminder$lifeExp)

## Call the hist() to view the life expectency distribution
hist(gapminder$lifeExp)
```

```{r}
## Call the table() function on the continent variable in the gapminder dataset to count the number of observations recorded for each continent.

table(gapminder$continent)

## Use a pie chart to show the distrubution of observations in africa

conts <- table(gapminder$continent)

pie(conts, labels =paste0(names(conts), ' ( ', round(prop.table(conts)*100), "%)"))

## Use a barplot to do the same but better.

barplot(conts, main="Frequency of Observations for each Continent")

```

From the table we can see the number of observations for each country, Africa has the highest frequency (n = 624) and Oceania with the lowest frequencey (n = 24)

Since the goal is to be able to work with complex data and break it down into component parts, I will focus on the continent of Oceania because if I want to work with variations between countries it will be simpler to work with 24 countries instead of 624, but ideally the same analysis could be used for any of the continents within the  data set.


## 1.4 Subset data and continue EDA
```{r}
## Subset data for the coninent Oceania, using pipes %>% and the filter() function.

df <- gapminder %>%
  filter(continent == "Oceania")

dfTest <- filter(gapminder, continent == "Oceania")

summary(df$country)

## ok so here we are saving a new dataframe however  ALL of the factor levels are saved into the new dataframe when ther are only 2 countries within Oceania. Its not necessarily that its missing (NA).  Why does table show all the factors for which there are no observations?
```

```{r}



```


```{r}

```


```{r}


```


```{r}


```

```{r}

```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}

```


```{r}

```


```{r}

```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
