---
title: "hw3-r4d-chapter7exercises"
author: "Irfan Ainuddin"
date: "9/20/2019"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
```
## 7.3.1 Visualizing distributions

```{r}

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))

diamonds %>% 
  count(cut)

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)

diamonds %>% 
  count(cut_width(carat, 0.5))

smaller <- diamonds %>% 
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)


ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

## 7.3.2 Typical Values

```{r}

ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

## 7.3.3 Unusual values

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))

unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
unusual
```

## 7.3.4 Exercises

```{r 7.3.4.1}
## Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

## inspect diamonds set
str(diamonds)
head(diamonds)

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = x))
## ranges from ~ 3.5 to 10

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = x), binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 15))


ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = y))
## ranges from ~ 3 to 10

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 15))


ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = z))
## ranges from ~ 2 to 6

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = z), binwidth = 0.5) +
  coord_cartesian(xlim = c(0, 10))




```
The x variables ranges from 4 to 10, the y variable ranges from 4 to 10 and the z varible ranges from 2.5 to 7.5. If you took at a diamond top down the width and length would be about even or squarish, and the height of the diamond would be longer or shorter so for those reasons i would guess that the x and y are width and length and the z value is the height, probably in milimeters

```{r 7.3.4.2}
## Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = price))

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 1000)

## ranges from o to 20,000
```

the distribution of the price of diamonds is positively skweed to the right, the bulk of the diamonds are on the lower end of the price range.

```{r 7.3.4.3}
## How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

unusualCarat <- diamonds %>% 
  filter(carat == .99) %>% 
  select(carat) %>%
  arrange(carat)
count(unusualCarat)

unusualCarat2 <- diamonds %>% 
  filter(carat == 1) %>% 
  select(carat) %>%
  arrange(carat)
count(unusualCarat2)
```

There are 23 diamonds recorded as .99 carat and 1558 recorded as 1 carat, the difference is probably caused by differences in the actual person who made the observation. Some people are specific some people generalize, or round up.

```{r 7.3.4.4}
## Compare and contrast coord_cartesian() vs xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = x), binwidth = 0.5) +
  coord_cartesian(xlim = c(5, 15))


ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = x), binwidth = 0.5) +
  xlim(5,15)
```

Using the cartesian xlim() method to zoomon the axies shows half of the bar and extends the range of the y axis to incoporate it, using ggplot's xlim() function the half bar does not show and the range of the y axis is reduced.  The ggplot xlim() method throws away any values outside of limit while the coord_cartesian() function retains the values and just limits the scope of the graph itself.
## 7.4

```{r}
diamonds2 <- diamonds %>% 
  filter(between(y, 3, 20))

head(diamonds2)

diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))

head(diamonds2)


ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()

## na.rm = TRUE removes issing data warning flag from above
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)

```


```{r}
## using mutate to capture useful missing values like in the case of a 0 flight time indicates a canceled

nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
## 
```
  
## 7.4.1 Exercises

```{r 7.4.1.1}

## What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference?


```
In a bar chart the missing values are captured as zero values and and displayed on the chart, in a histogram the missing values seem to be omitted and not displayed, or perhaps since the histogram is binning the values it is captured in the first value bin and counted as having a value?




```{r 7.4.1.2}
## What does na.rm = TRUE do in mean() and sum()?
mean(diamonds$x)
mean(diamonds$x, na.rm = TRUE)


sum(diamonds$x)
sum(diamonds$x, na.rm = TRUE)
```


I do not seem to see an effect of setting na.rm = TRUE?

## 7.5.1

```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))

ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()

ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip()
```

## 7.5.1.1 Exercises

```{r 7.5.1.1.1}
## Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights.
flights <- nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  )
str(flights)

## base graph
  ggplot(flights,mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)

## lets improve by....


```

```{r 7.5.1.1.2}
## What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?


ggplot(data = diamonds2) +
  geom_boxplot(mapping = aes(x = reorder(clarity, color, FUN = carat), y = price))
```

```{r 7.5.1.1.3}

## Install the ggstance package, and create a horizontal boxplot. How does this compare to using coord_flip()?


```

```{r 7.5.1.1.4}

## One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using geom_lv() to display the distribution of price vs cut. What do you learn? How do you interpret the plots?
```


```{r 7.5.1.1.5}
## Compare and contrast geom_violin() with a facetted geom_histogram(), or a coloured geom_freqpoly(). What are the pros and cons of each method?

```

```{r 7.5.1.1.6}

##If you have a small dataset, it’s sometimes useful to use geom_jitter() to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to geom_jitter(). List them and briefly describe what each one does.


```

## 7.5.2.1 Exercises

```{r 7.5.2.1.1}

## How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?

```

```{r 7.5.2.1.2}
## Use geom_tile() together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?


```

```{r 7.5.2.1.3}
## Why is it slightly better to use aes(x = color, y = cut) rather than aes(x = cut, y = color) in the example above?


```

## 7.5.3.1 Exercises

```{r 7.5.3.1.1}
## Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs cut_number()? How does that impact a visualisation of the 2d distribution of carat and price?

```

```{r}
## Visualise the distribution of carat, partitioned by price.
```

```{r}
## How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?


```

```{r}
## Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately.



```


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
